A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-02 shanghai.tex

In Part 2 we use Theorem~\ref{thm:snake} to make a sheaf--function dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$; the commutative
group scheme $\GN{T}_n$ plays the r\^{o}le of $G$.  Theorem~\ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$:
\[
\QCiso{\GN{T}_n} 
\iso
\Hom_{< n}(T(K),\EEx).
\]  
We also describe a pro-group scheme $\GN{T}$ simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application}. 
%\begin{enumerate}
%\item
Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of Lusztig's theory of characters sheaves to the study of characters of admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori. 
%\item
DeBacker and Reeder have recently applied Deligne-Lusztig representations to part of the
local Langlands correspondence. % \cite{reeder-debacker:09a}.  
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
We hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups. 
%\item
The scope of the Langlands program has grown explosively in the past several decades, and
different parts of it apply to different kinds of representations.  In particular, the classic local Langlands
correspondence addresses complex (or $\ell$-adic) representations of real and $p$-adic groups, while
the geometric Langlands program aims at representations of groups in characteristic $p$.  The fact that
characteristic $p$ group schemes arise as the space underlying quasicharacter sheaves, regardless of
the characteristic of $K$, provides a potential bridge allowing the methods of the geometric Langlands
program to be applied to the classic context. 
%\item
Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
\[
\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \Hom(T(K), \EEx).
\]
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.
%\end{enumerate}

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
%\cite{boyarshenko-drinfeld:motivated}*{\S 1.8}.  
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is of course due to Lusztig. It is applied to certain perverse
sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with a Galois action.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the sheaf--function dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{\Fq}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

>>>>

In Part 2 we use Theorem~\ref{thm:snake} to make a sheaf--function dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$; the commutative
group scheme $\GN{T}_n$ plays the r\^{o}le of $G$, above.  Theorem~\ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$:
\[
\QCiso{\GN{T}_n} 
\iso
\Hom_{< n}(T(K),\EEx).
\]  
We also describe a pro-group scheme $\GN{T}$ simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application} here.

Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of Lusztig's theory of characters sheaves to the study of characters of admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori.

DeBacker and Reeder have recently applied Deligne-Lusztig representations to part of the
local Langlands correspondence. % \cite{reeder-debacker:09a}.  
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
We hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups.

The scope of the Langlands program has grown explosively in the past several decades, and
different parts of it apply to different kinds of representations.  In particular, the classic local Langlands
correspondence addresses complex (or $\ell$-adic) representations of real and $p$-adic groups, while
the geometric Langlands program aims at representations of groups in characteristic $p$.  The fact that
characteristic $p$ group schemes arise as the space underlying quasicharacter sheaves, regardless of
the characteristic of $K$, provides a potential bridge allowing the methods of the geometric Langlands
program to be applied to the classic context.

Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
$
\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \Hom(T(K), \EEx)
$. 
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
%\cite{boyarshenko-drinfeld:motivated}*{\S 1.8}.  
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of certain commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is of course due to Lusztig. It is applied to certain perverse
sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the sheaf--function dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{\Fq}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.
